---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(unbiasedgoodness)
```

A model chosen by a data-driven method will often look like it fits well, because the method chose it for fitting well. This package tests the fit of a selected linear model against a data-driven enlargement of it: if the selected model has $s$ variables, the alternative adds variables until it has about $\lceil m s + b \rceil$.

We simulate a sparse high-dimensional regression with [hdi::rXb()]; the true support is the first `s0` columns.

```{r}
set.seed(1)
n <- 100
p <- 60
s0 <- 3
sim_data <- hdi::rXb(n, p, s0, xtype = "toeplitz", x.par = 1 / 3, verbose = FALSE)
x <- sim_data$x
y <- as.numeric(x %*% sim_data$beta + rnorm(n))
```

## Selective F test after forward stepwise

`gof_groupfs()` selects a model by forward stepwise with a penalized stopping rule and returns a p-value that is valid conditional on the selected and enlarged models.

```{r}
gof_groupfs(x, y, m = 1.2, b = 5)
```

## Residual prediction tests after the lasso

`gof_RPtest()` selects a model with the cross-validated lasso and applies the residual prediction tests of Shah and Bühlmann (2018), implemented in the **RPtests** package. These p-values are not adjusted for selection; the articles on the package website study their selective error rates by simulation.

```{r}
out <- gof_RPtest(x, y, m = 1.2, b = 5, B = 19L)
out[c("selected_support", "alt_support")]
unlist(out[startsWith(names(out), "pval")])
```

## Simulation

`simulate_gof()` repeats data generation, selection and testing, returning one row per iteration. The column `null` records whether the selected model contained the true support.

```{r}
sims <- simulate_gof(4, m = 1.2, b = 5, n = 100, p = 60, s0 = 3,
                     method = "groupfs", seed = 1)
sims[, c("s_null", "s_alt", "null", "pval_selective", "pval_naive")]
```
