---
title: "RP tests"
bibliography: ../bibliography.bib
link-citations: TRUE
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Philosophy based on cross-validation

In chapter 7 of @ESL2009 the authors emphasize cross-validation estimates the expected prediction error (or generalization error) of a *method*. This is in contrast to estimating such errors for a *specific model*. There is a line of work that makes use of cross-validation and inherits its inferential focus on a method rather than a specific model chosen by that method [@wasserman2009; @meinshausen2009; @meinshausen2010; @shah2013] including the focus of this vignette which was proposed by @shah2018. This paper and its `R` implementation [-@rRPtests] roughly attempt to answer the question:

[**Does this method (typically) result in models with a good fit, or does it fail to capture signal that an alternate method could (typically) detect?**]{.smallcaps}

If the first method does fit well, then the residuals from that method should (typically) look like pure noise. On the other hand, if an alternate method can predict the residuals with good accuracy (better than noise) that would be evidence the first method is failing to capture some signal. Hence the name **RPtests**, based on **r**esidual **p**rediction. The exact details depend on several (groups of) decisions:

1.  Probability modeling assumptions (foundational, and we will not question these here).$$
    \mathbf{y} = \mathbf{X} \beta + \sigma \epsilon
    $$

2.  The first method, or null method. We want to know if this method is good enough. *We will focus on using the lasso [-@tibshirani1996] to select variables* in a (potentially high-dimensional) linear model. (The high level idea could be applied using other methods as well but that would require changing some other details in the implementation).

3.  One or more secondary or alternate methods that could be chosen depending on what type of model misspecification we are concerned about. *We will focus on whether we should include more predictor variables*, but another possibility is to search for non-linear relationships. (An alternate method could be a hybrid or ensemble of several methods, but we will not explore this here).

4.  Analysis/pipeline choices including how to tune each method, estimating the variance of the noise, number of bootstrap samples, and so on. (We will also not question these here).

We'll focus on one of the main cases investigated in [@shah2018]: testing a linear model against an alternative that includes additional predictors. To determine if the alternative predicts the residuals significantly better than noise the paper uses a parametric-residual bootstrap described below, but first we consider an important point about the dimension of the first (null) model.

## Low vs high-dimensions

In the low dimensional linear model case, the scaled residuals

$$
\hat{\mathbf{R}} = \frac{(\mathbf{I} - \mathbf{P})\mathbf{y}}{\|(\mathbf{I} - \mathbf{P})\mathbf{y}\|_2}
$$

are a pivotal quantity under the null hypothesis that the model fits (and $\mathbf P = \mathbf{X}(\mathbf{X}^T \mathbf{X})^{-1} \mathbf{X}^T$ is the projection or hat matrix). In this case the distribution of the residuals does not depend on any unknown parameters, so we can use that distribution for valid hypothesis tests or other inferences. This is the basis of the standard $F$-test for goodness-of-fit, the one reported at the end of a `summary(lm(y ~ x))`.

```{r }
set.seed(1)
x <- runif(50)
y <- x + rnorm(50)
summary(lm(y ~ x))
```

However, this does not work in the high-dimensional case, so [-@shah2018] use a combination of previous lasso theory and new technical results to prove the residuals from lasso can be used in a similar way (with slight modifications) to construct tests. The theory relies on fairly standard but strong conditions sufficient for the lasso to estimate the support and signs of the true $\beta$ with high probability, but (as is typical in lasso theory literature) simulations show good empirical performance under more general settings where the strong sufficient conditions are relaxed.

We briefly note that @jankova2020 apply the same basic idea of predicting residuals to high-dimensional *generalized* linear models by using the Pearson residuals

$$
\hat R_i = \frac{Y_i - \mu(x_i^T \hat \beta)}{\sqrt{V(x_i^T \hat \beta)}}
$$

instead, where $\mu$ and $V$ are the conditional mean and variance functions, respectively. However, beyond the basic idea there is much additional work involved in this other case so we do not consider it more here.

### Lasso details

To calibrate the test, we need a distribution to compare the accuracy measure (typically) achieved by a null model. In the high-dimensional case the null model is assumed to be selected by lasso with cross-validation. The authors propose iterating the whole procedure multiple times to average over the randomness induced by cross-validation itself. Note that this makes more sense when our inferential focus is the *method* and not a *specific model* chosen by the method. Below is a restatement of Algorithm 1 in [-@shah2018], the parametric-residual bootstrap procedure for estimating a null distribution. For $b = 1, \ldots, B$:

1.  Replace $\mathbf{y}$ with $\mathbf{\hat y}$ from the original null model plus simulated residuals $\mathbf{\zeta}^{(b)} \sim N(0, \mathbf{I}_{n \times n})$. (Simulating a new response from the null model).\
2.  Fit sqrt-lasso [-@belloni2011] using the above simulated $\mathbf{y}^{(b)}$ and subtract new fit $\mathbf{\hat y}^{(b)}$. (Simulating residuals achieved by alternative method).\
3.  Normalize simulated residuals and compute some accuracy measure, such as RSS. (Obtain bootstrap distribution for testing).\

Note that the sqrt-lasso method can be replaced with other fitting procedures if we are interested in different alternatives, for example by using a tree based method if we are interested in predictive interactions.

Exercise: why not just predict normalized residuals $(\mathbf{y} - \mathbf{\hat y})/\hat \sigma$ and compare to accuracy in predicting $\mathbf{\zeta}^{(b)}$? (Consider dimensions).

### `RPtest` code explanation

Below we do not consider special cases, for example when a user-specified `noise_matrix` or `beta_est` overrides default behavior. We assume the function is called as `RPtest(x, y, resid_type, test = "group", x_alt)` with no other options changed.

**Note**: with `test = "group"` the `RPfunction` will be set to `lasso_test` regardless of the dimension of the null or alternative.

#### `RPtest`

When `resid_type = "Lasso"` (the default unless otherwise specified) the following functions will be called.

First, `cv.glmnet` is run several times (in parallel) to get an estimate `beta_est`, unless one is supplied. (A supplied `beta_est` is ignored when `resid_type = "OLS"`.) The estimated signal is `sig_est <- x %*% beta_est`. 

Next, `comp_sigma_est` is called with this `beta_est` to compute `sqrt(MSE)`. This is the noise level used by `rnorm`, the default `rand_gen`, to generate noise terms for simulating residuals.

Now simulated residuals `resid_sim` are computed by calling `resid_gen_lasso` with the above estimates and noise terms.

#### `resid_gen_lasso`

In parallel, calls `resid_lasso` which fits square-root lasso and returns residuals. In the implementation of `sqrt_lasso`, if `x` is not high-dimensional then you will see warnings saying `Smallest lambda chosen by sqrt_lasso`. 

If the type is `"OLS"` then instead of `sqrt_lasso` residuals are computed the ordinary way. 

#### Orthogonalizing `x_alt`

Based on `resid_type`, this part either uses `sqrt_lasso` or OLS to residualize `x_alt`, subtracting away predictions based on fits to `x_null`. By default `x_null = x`, the input to `RPtest`.

**Note**: At this point, if the input `x_alt` is a subset of `x`, i.e. using the common nested model structure of $F$-tests, this residualization will practically project `x_alt <- 0` in the `"OLS"` case or close to it in the `"Lasso"` case.

#### The default `RPfunction` is `lasso_test` (when `test = "group"`)

This is applied on both the original and simulated residuals using the above orthogonalized `x_alt`, and will use `glmnet(x_alt, resid[_sim])`. As a test statistic it computes the average MSE over the full solution path using the same values of `lambda` as fit to the original residuals.

#### Summary

After projecting `x_alt` (approximately, in the Lasso case) orthogonal to `x`, it is used with `glmnet` to predict the original residuals `resid` and simulated ones `resid_sim`, and the MSE over the solution paths are compared. If this `x_alt` predicts the original residuals better than those simulated from the null then it's evidence there is signal remaining in the (original) residuals. To use this as a goodness-of-fit test we therefore make `x_alt` include some additional variables which were not selected in the original fit.

## Strategy for goodness-of-fit test

The original model selection algorithm chooses some number of variables, let's call this `s0`. To test for the presence of signal variables that were not included in this original fit, we choose a larger sparsity level `m * s0 + b` with `m >= 1` and `b > 0`. We use the model selection algorithm to include more variables up to this level of sparsity, and we test the significance of these new additional variables as a way of checking the fit of the original, smaller model.

## Simulations and examples

```{r setup, message=FALSE}
library(unbiasedgoodness)
library(dplyr)
library(tidyr)
library(ggplot2)
```

We continue focusing on the high-dimensional linear model and as an alternative to a selected model we consider whether more variables should be included. Since @shah2018 are not trying to do selective inference there is no model selection involved in choosing the hypotheses in their simulations. Type 1 error control is shown in simulations set up for the null to be true in all instances, rather than doing model selection and subsetting to instances where the null holds (as in selective inference). In contrast to their simulations, we are interested in the selective (conditional) error control rates, as we explore in the simulations below.

### One data generation example

```{r}
set.seed(1)
n <- 100
p <- 200
s0 <- 5
sim_data <- hdi::rXb(n, p, s0, xtype = "equi.corr", x.par = 1 / 20, verbose = FALSE)
x <- sim_data$x
beta <- sim_data$beta
y <- as.numeric(x %*% beta + rnorm(n))
# True support
which(beta != 0)
```

```{r}
gof_RPtest(x, y)
```

The p-values are

- `pval_OLS`, `pval_lasso`: the added variables (`alt_support`) as `x_alt`, with OLS or square-root lasso residuals of the selected model;
- `pval_lasso_beta`: as `pval_lasso`, but the residual bootstrap uses the lasso coefficients of the selected model as `beta_est` instead of re-running `cv.glmnet` on the selected variables;
- `pval_full`, `pval_full_lasso`: all unselected variables as `x_alt`, i.e. the original goodness-of-fit test of @shah2018 applied to the selected model.

### One simulation iteration

```{r}
set.seed(1)
m <- 1.2
b <- 10
simulate_gof_instance(m, b, n, p, s0)
```

The column `null` is `TRUE` when the selected model contains all `s0` true variables, so that the goodness-of-fit null hypothesis holds for the selected model.

### Iterate many times

The results below were computed by `data-raw/sim_rptest.R` and are loaded from the installed package, since the simulation takes a while.

```{r}
output <- read.csv(system.file("extdata", "sim_rptest.csv",
                               package = "unbiasedgoodness"))
nrow(output)
```

The settings are $n = 100$, $p = 200$, $s_0 = 5$, a Toeplitz design with correlation $(1/3)^{|i-j|}$, nonzero coefficients drawn from $U[-2, 2]$, unit noise variance, $m = 1.2$, $b = 10$ and $B = 49$ bootstrap samples per test.

### Distributions of $p$-values

```{r glmnet-RPtest-pval-plots, fig.width = 8, fig.height = 4}
pvals <- output |>
  pivot_longer(starts_with("pval"), names_to = "type", values_to = "pvalue") |>
  mutate(null = if_else(null, "selected model correct", "selected model misses signal"))

ggplot(pvals, aes(pvalue, colour = type)) +
  stat_ecdf() +
  geom_abline(slope = 1, linetype = "dashed") +
  facet_wrap(~ null) +
  coord_equal() +
  labs(x = "p-value", y = "empirical cdf", colour = NULL)
```

### Selective error control

Rejection rates at level 0.05, with Monte Carlo standard errors:

```{r}
rejection_rates <- pvals |>
  group_by(null, type) |>
  summarize(instances = n(),
            rejection_rate = mean(pvalue < 0.05),
            mc_se = sqrt(rejection_rate * (1 - rejection_rate) / instances),
            .groups = "drop")
knitr::kable(rejection_rates, digits = 3)
```

Within `selected model correct` the rejection rate estimates the selective type 1 error $P(\text{reject} \mid \text{selected model contains the true support})$. This averages over the models selected in that event; it does not show error control conditional on each particular selected model. Within `selected model misses signal` it estimates power, but note that coefficients drawn from $U[-2, 2]$ are often very small, so many of these instances are close to the null.

## Model selection success rates

These diagnostics give some indication of the difficulty of lasso model selection for this data generating process.

```{r}
output |>
  summarize(null = mean(null),
            alt = mean(alt),
            s_null = mean(s_null),
            s_alt = mean(s_alt))
```

```{r, fig.width = 8, fig.height = 3}
output |>
  select(null_intersection, alt_intersection) |>
  pivot_longer(everything()) |>
  ggplot(aes(value)) +
  geom_bar() +
  facet_wrap(~ name) +
  labs(x = "number of true variables included")
```

```{r, fig.width = 8, fig.height = 3, message = FALSE}
output |>
  select(null_intersection, max_corr, min_beta) |>
  pivot_longer(c(max_corr, min_beta)) |>
  ggplot(aes(value, null_intersection)) +
  geom_jitter(height = 0.1, alpha = 0.3) +
  geom_smooth(method = "lm") +
  facet_wrap(~ name, scales = "free_x")
```

### Does RPtests control selective error?

```{r, include = FALSE}
rate <- function(col) {
  keep <- rejection_rates$null == "selected model correct" & rejection_rates$type == col
  rejection_rates$rejection_rate[keep]
}
```

**RPtests** was not designed to control selective errors, and in this simulation it does not when the alternative is chosen by the lasso. Among the `r sum(output$null)` iterations in which the selected model contains the true support, the tests of the added variables reject at level 0.05 with frequency `r round(rate("pval_OLS"), 2)` (`pval_OLS`) and `r round(rate("pval_lasso"), 2)` (`pval_lasso`). When the selected model is correct the added variables are pure noise, but they are the noise variables most correlated with the residuals of the selected model, since that is how the lasso chose them. A test of whether they predict those residuals that ignores this choice is anti-conservative. (An earlier version of this article suggested the opposite conclusion. Its simulation had silently used `b = 5` instead of `b = 10`, but rerunning the original 2022 code with either value, on 120 iterations each, also rejected in well over half of the iterations in which the selected model was correct.)

The tests that use *all* unselected variables as the alternative (`pval_full`, `pval_full_lasso`) do not involve a data-driven alternative, and their rejection rates here are at or below the nominal level, at the cost of having essentially no power against the alternatives in this simulation.

These results concern one data generating process and one tuning of the procedure.

## References
