---
title: "Selective goodness-of-fit after forward stepwise"
bibliography: ../bibliography.bib
link-citations: TRUE
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Philosophy based on conditional inference

[**Does this specific selected model have a good fit, or does it fail to capture signal that an alternate model could detect?**]{.smallcaps}

In contrast to the residual prediction approach (see the RP tests article), which asks whether a *method* typically produces models that fit, the selective inference approach [@fithian2014] asks about the *specific model* that was selected. A test of a hypothesis chosen by looking at the data is calibrated using the distribution of the test statistic conditional on the selection event, i.e. on having chosen that hypothesis.

## Low vs high-dimensions

The standard $F$-test for goodness-of-fit for an *a priori* fixed model compares its fit to that of a larger fixed model. The overall $F$-test reported at the end of a `summary(lm(y ~ x))` is the special case comparing an intercept-only model to the full model:

```{r}
set.seed(1)
x <- runif(50)
y <- x + rnorm(50)
summary(lm(y ~ x))
```

To check the fit of a selected model $M$ we could test it against the model containing all $p$ predictors. When $p > n$ this is impossible, since the full model is saturated and the residual degrees of freedom $n - 1 - p$ are negative. Instead we test the selected model against a data-driven enlargement of it.

## The procedure

`gof_groupfs()` uses forward stepwise with groups [@loftus2015] as the model selection method.

1. **Select.** Run forward stepwise. At each step add the group that minimizes the penalized criterion $n \log(\mathrm{RSS}/n) + k \cdot \mathrm{df}$ (or $\mathrm{RSS}/\sigma^2 + k \cdot \mathrm{df}$ if $\sigma$ is known). Stop when the criterion has increased `aicstop` times in a row. The first $s$ groups form the selected (null) model $M_0$.
2. **Enlarge.** Continue forward stepwise to $M = \max(\lceil m s + b \rceil, s + \texttt{aicstop})$ groups, the alternative model $M_1 \supset M_0$. $M$ is capped so that $M_1$ is not saturated.
3. **Test.** With $P_0$ and $P_1$ the projections onto the column spaces of $M_0$ and $M_1$ (after centering), use the usual statistic
$$
T_F = \frac{\|(P_1 - P_0) y\|^2 / \mathrm{df}_1}{\|(I - P_1) y\|^2 / \mathrm{df}_2},
\qquad \mathrm{df}_1 = \mathrm{rank}(P_1) - \mathrm{rank}(P_0), \quad
\mathrm{df}_2 = n - 1 - \mathrm{rank}(P_1),
$$
for $H_0: \mu \in \mathrm{span}(1, X_{M_0})$.

### Conditioning

Write $y = Z + R\,(g_1(T) V_\delta + g_2(T) V_2)$ where $Z = P_0 y$, $R = \|y - Z\|$, $V_\delta$ and $V_2$ are the unit vectors in the directions of $(P_1 - P_0) y$ and $(I - P_1) y$, and $g_1(t) = \sqrt{Ct/(1 + Ct)}$, $g_2(t) = 1/\sqrt{1 + Ct}$ with $C = \mathrm{df}_1/\mathrm{df}_2$. Under $H_0$ with Gaussian errors, $T_F$ is independent of $(Z, R, V_\delta, V_2)$ and has an $F_{\mathrm{df}_1, \mathrm{df}_2}$ distribution.

Holding $(Z, R, V_\delta, V_2)$ fixed, each comparison made by forward stepwise (every candidate group at every step up to $M$) and the sign of each change in the penalized criterion (up to step $s + \texttt{aicstop}$) is a constraint on $t$, the value of $T_F$. The set of $t$ satisfying all of them is a union of intervals $E$, and the selective p-value is
$$
P(F_{\mathrm{df}_1, \mathrm{df}_2} \geq T_F \mid F_{\mathrm{df}_1, \mathrm{df}_2} \in E).
$$
Because conditioning on the whole forward stepwise path and the signs of all the criterion differences is finer than conditioning on the selected pair $(M_0, M_1)$, the resulting test controls the type 1 error conditional on $(M_0, M_1)$, at some cost in power. When $\sigma$ is known the same construction is used with a truncated $\chi$ statistic.

## Example

```{r setup, message=FALSE}
library(unbiasedgoodness)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r}
set.seed(1)
n <- 100
p <- 200
s0 <- 5
sim_data <- rXb(n, p, s0, xtype = "equi.corr", x.par = 1 / 20, verbose = FALSE)
x <- sim_data$x
beta <- sim_data$beta
y <- as.numeric(x %*% beta + rnorm(n))
# True support
which(beta != 0)
```

```{r}
fit <- gof_groupfs(x, y, m = 1.2, b = 10)
fit
```

`p_naive` is the p-value of the same $F$ statistic ignoring selection. The truncation region printed as `trunc_min` and `trunc_max` is typically short, which is why the selective test can have low power.

## Simulations

The results below were computed by `data-raw/sim_groupfs.R` and are loaded from the installed package.

```{r}
output <- read.csv(system.file("extdata", "sim_groupfs.csv",
                               package = "unbiasedgoodness"))
nrow(output)
```

The settings are $n = 100$, $p = 200$, $s_0 = 5$, a Toeplitz design with correlation $(1/3)^{|i-j|}$, nonzero coefficients drawn from $U[-2, 2]$, unit noise variance, $k = \log p$, $m = 1.2$ and $b = 10$.

Of 1000 iterations, `r 1000 - nrow(output)` were dropped because the stopping rule never stopped: with $\sigma$ unknown and $p > n$ the criterion $n \log(\mathrm{RSS}/n) + k \cdot \mathrm{df}$ can keep decreasing until the model is saturated, leaving nothing to test against. The rates below are conditional on the stopping rule having stopped.

### Distributions of $p$-values

```{r groupfs-pval-plots, fig.width = 8, fig.height = 4}
pvals <- output |>
  pivot_longer(starts_with("pval"), names_to = "type", values_to = "pvalue") |>
  mutate(null = if_else(null, "selected model correct", "selected model misses signal"))

ggplot(pvals, aes(pvalue, colour = type)) +
  stat_ecdf() +
  geom_abline(slope = 1, linetype = "dashed") +
  facet_wrap(~ null) +
  coord_equal() +
  labs(x = "p-value", y = "empirical cdf", colour = NULL)
```

### Selective error control

Rejection rates at level 0.05, with Monte Carlo standard errors:

```{r}
pvals |>
  group_by(null, type) |>
  summarize(instances = n(),
            rejection_rate = mean(pvalue < 0.05),
            mc_se = sqrt(rejection_rate * (1 - rejection_rate) / instances),
            .groups = "drop") |>
  knitr::kable(digits = 3)
```

Within `selected model correct` the rejection rate estimates $P(\text{reject} \mid \text{selected model contains the true support})$. The theory above gives the stronger guarantee of error control conditional on each selected pair $(M_0, M_1)$. Within `selected model misses signal` the rejection rate estimates power. Coefficients drawn from $U[-2, 2]$ are often very small, so many of these instances are close to the null.

In this design the selective test's error rate is at or below 0.05 when the selected model is correct, but it has almost no power when the selected model misses signal: its rejection rate there is barely above its rate under the null. The naive $F$ test rejects in every iteration, whether or not the selected model is correct, so it is useless as a diagnostic here. The selected models are large (on average `r round(mean(output$s_null), 1)` variables for $s_0 = 5$), so the added variables are tested with few residual degrees of freedom and a short truncation region.

## Model selection success rates

```{r}
output |>
  summarize(null = mean(null),
            alt = mean(alt),
            s_null = mean(s_null),
            s_alt = mean(s_alt))
```

```{r, fig.width = 8, fig.height = 3}
output |>
  select(null_intersection, alt_intersection) |>
  pivot_longer(everything()) |>
  ggplot(aes(value)) +
  geom_bar() +
  facet_wrap(~ name) +
  labs(x = "number of true variables included")
```

## References
